tstrait: a quantitative trait simulator for ancestral recombination graphs

### 2023-11-27 paper.tex
Pass

In Fig.~\ref{fig:tree-illustration} we show an example
tree for three individuals.
Because these individuals are diploids, each is associated
with two nodes in the tree (highlighted by colour).
Ancestral nodes are not associated with individuals here,
but in general the ARG may be embedded in a multigenerational
pedigree, where some internal nodes would be associated
with individuals.
In the example of
Fig.~\ref{fig:tree-illustration}, T is chosen as the causal
allele and $\beta=0.05$, and
therefore all nodes descending from \textsf{i}
have genetic value $0.05$, except \textsf{e} which
has zero because of the back-mutation to A.
Following the standard practise in GWAS~\citep{uffelmann2021},
we assume the additive model such that the overall
genetic value of an individual is the sum of its
nodes' genetic values.
Given these per-individual genetic values, the final phenotype
is then generated by adding some environmental noise.
This noise is simulated from a normal distribution with mean zero
and variance of $V_G(1-h^2)/{h^2}$,
where $V_G$ is the variance of the individual genetic values
and $h^2$ is the narrow-sense heritability provided as input by the user.

\subsection{Interface}
\texttt{Tstrait} is a Python library, building on the \texttt{tskit}
ARG library~\citep{ralph2020,wong2023general} and the rich
Python data-science ecosystem~\citep{numpy}.
Simulating a phenotype for an ARG with default parameter
values requires only a few lines of code:
% NOTE this listing assumes num_causal has been moved, or removed.
% We need to check whether this actually works in the final version.
\begin{lstlisting}[language=Python]
import tstrait as tst
model = tst.trait_model('normal', mean=0, var=1)
result = tst.sim_phenotype(arg, model)
\end{lstlisting}
We first create \texttt{model}, which represents the distribution
from which effect sizes are drawn. Five commonly used
univariate distributions are supported, along with the
multivariate normal distribution to model pleiotropic traits.
Given this model, we can then simulate phenotypes for the individuals
in an ARG (as a  \texttt{tskit TreeSequence}) using the
\texttt{sim\_phenotype} function.
The user can either specify a number of causal sites to be chosen randomly
along the genome (one, by default), or can directly provide the causal
sites as input. Combined with the detailed information about
mutations recorded in a \texttt{tskit} ARG, this provides the ability
to model population and haplotype specific effects.
The return value \texttt{result} is an object encapsulating
two Pandas dataframes~\citep{mckinney2010data}: one describing the simulated
effect sizes and the other describing the genetic values,
environmental noise and final phenotypes for each individual.
The simulation results can then be efficiently and conveniently
processed by a range of familiar Python data science tools.

As well as this convenient single-function interface,
\texttt{tstrait} provides modular building blocks for power-users
and potential integration with other tools that generate
traits on an ARG. The \texttt{sim\_trait} function
simulates effect sizes for an input ARG, and returns a data frame
describing the causal alleles and effect sizes.
This dataframe can then be passed to the \texttt{genetic\_values}
function, which calculates the genetic values for each
node, and accumulates them by individual
(Fig~\ref{fig:tree-illustration}). Finally, the \texttt{sim\_env}
function takes these per-individual genetic values
and adds some simulated environmental noise to produce the final
phenotypes.

A major benefit of this modular architecture is the flexibility it offers
users. Because the causal sites and effect sizes are specified in a
simple tabular format, users can easily develop their own approach
to simulating these values. Alternatively, other simulators
such as SLiM~\citep{haller2023} that generate effect sizes and
causal mutations during the progress of a forwards-time simulation
(to model the effect of selection, for example) could output these
values to a CSV or similar file. The modular architecture  and simple
input data formats are specifically intended to facilitate such
interoperability.

@@

\subsection{Performance}
\texttt{Tstrait} is very efficient, and can be applied to datasets at the
largest scales.
Supplementary Fig~\ref{fig:time}A shows how trait simulation time scales
with number of individuals, for sequences of length 50, 100 and 200Mb.
For each sample size and sequence length,
we first simulate an ARG under a simple coalescent model with
human-like parameters using~\texttt{stdpopsim}~\citep{adrion2020}.
We then run a quantitative trait simulation
with 1000 causal sites on this ARG using \texttt{tstrait},
and report the mean CPU time over 10 independent replicates.
Supplementary Fig~\ref{fig:time}B shows that RAM usage is very
modest for all these simulations.
A comparison with existing methods that require reading
in genetic sequences in formats such as
VCF~\citep[e.g.][]{meyer2018,fernandes2020} is not feasible,
as the VCFs involved are very large, and these tools assume that
the entire dataset fits in RAM.
To emphasise this point, we also generated
quantitative traits for the simulations based on
the French-Canadian pedigree~\citet{anderson2023}, discussed in
the introduction. It took 60 seconds and XXX GB of RAM to
simulate a trait with 1000 causal sites using \texttt{tstrait}.
Using one byte per genotype, the full variant matrix of this
dataset would require 140TB of RAM.
% If we have space, mention the real-data sims

% Figure \ref{fig:time} (B) shows simulation time of \texttt{tstrait} with more
% realistic data. For example, \texttt{tstrait} is used to simulate traits from
% 2.7 million individuals in the French Canadian dataset, which is the largest
% known simulated genetic dataset \citep{anderson2023}. It took 60 seconds to
% simulate a trait with 1000 causal sites, making \texttt{tstrait} to be
% practical to simulate quantitative traits from this largest ever simulated
% dataset. The inferred ARG from the 1000 Genomes project \citep{kelleher2019} is
% also used to measure the computational performance.

>>>>

In Fig.~\ref{fig:tree-illustration} we show an example
tree for three individuals.
Because these individuals are diploids, each is associated
with two nodes in the tree (highlighted by colour).
Ancestral nodes are not associated with individuals here,
but in general an ARG may be embedded in a multigenerational
pedigree, where some internal nodes would be associated
with individuals.
In the example of
Fig.~\ref{fig:tree-illustration}, T is chosen as the causal
allele and $\beta=0.05$, and
therefore all nodes descending from \textsf{i}
have genetic value $0.05$, except \textsf{e} which
has zero because of the back-mutation to A.
Following the standard practise in GWAS~\citep{uffelmann2021},
we assume the additive model such that the overall
genetic value of an individual is the sum of its
nodes' genetic values.
Given these per-individual genetic values, the final phenotype
is then generated by adding some environmental noise.
This noise is simulated from a normal distribution with mean zero
and variance of $V_G(1-h^2)/{h^2}$,
where $V_G$ is the variance of the individual genetic values
and $h^2$ is the narrow-sense heritability provided as input by the user.

\subsection{Interface}
\texttt{Tstrait} is a Python library, building on the \texttt{tskit}
ARG toolkit~\citep{ralph2020,wong2023general} and the rich
Python data-science ecosystem~\citep{numpy}.
Simulating a phenotype for an ARG with default parameter
values requires only a few lines of code:
% NOTE this listing assumes num_causal has been moved, or removed.
% We need to check whether this actually works in the final version.
\begin{lstlisting}[language=Python]
import tstrait as tst
model = tst.trait_model('normal', mean=0, var=1)
result = tst.sim_phenotype(arg, model)
\end{lstlisting}
We first create \texttt{model}, which represents the distribution
from which effect sizes are drawn. Five commonly used
univariate distributions are supported, along with the
multivariate normal distribution to model pleiotropic traits.
Given this model, we can then simulate phenotypes for the individuals
in an ARG (as a  \texttt{tskit TreeSequence}) using the
\texttt{sim\_phenotype} function.
The user can either specify a number of causal sites to be chosen randomly
along the genome (one, by default), or can directly provide the causal
sites as input. Combined with the detailed information about
mutations recorded in a \texttt{tskit} ARG, explicitly specifying causal
sites allows us to model many different types of trait, for example
those associated with mutations arising in a particular population
or time interval.
The return value \texttt{result} is an object encapsulating
two Pandas dataframes~\citep{mckinney2010data}: one describing the simulated
effect sizes and the other describing the genetic values,
environmental noise and final phenotypes for each individual.
The simulation results can then be efficiently and conveniently
processed using familiar Python data science tools.

As well as this convenient single-function interface,
\texttt{tstrait} provides modular building blocks for power-users
and to facilitate integration with other tools that generate
traits on an ARG. The \texttt{sim\_trait} function
simulates effect sizes for an input ARG, and returns a data frame
describing the causal sites, alleles and effect sizes.
This dataframe can then be passed to the \texttt{genetic\_values}
function, which calculates the genetic values for each
node, and accumulates them by individual
(Fig~\ref{fig:tree-illustration}). Finally, the \texttt{sim\_env}
function takes these per-individual genetic values
and adds some simulated environmental noise to produce the final
phenotypes.

A major benefit of this modular architecture is the flexibility it offers
users. Because the causal sites and effect sizes are specified in a
simple tabular format, users can easily develop their own approach
to simulating these values. Alternatively, other simulators
such as SLiM~\citep{haller2023} that generate effect sizes and
causal mutations during the progress of a forwards-time simulation
could output these
values to a CSV or similar file. The modular architecture  and simple
input data formats are specifically intended to facilitate such
interoperability.

@@

\subsection{Performance}
\texttt{Tstrait} is very efficient, and can be applied to datasets at the
largest scales.
Supplementary Fig~\ref{fig:time}A shows how trait simulation time scales
with number of individuals, for sequences of length 50, 100 and 200Mb.
For each sample size and sequence length,
we first simulate an ARG under a simple coalescent model with
human-like parameters using~\texttt{stdpopsim}~\citep{adrion2020}.
We then run a quantitative trait simulation
with 1000 causal sites on this ARG using \texttt{tstrait},
and report the mean CPU time over 10 independent replicates.
Supplementary Fig~\ref{fig:time}B shows that RAM usage is very
modest for all these simulations.
To demonstrate its scalability we  applied \texttt{tstrait} to the
French-Canadian simulations~\cite{anderson2023} of chromosome 9
discussed in the introduction.
% TODO redo this for 100 causal sites @daiki.
It took 6 seconds and XXX GB of RAM to
simulate a trait with 100 causal sites using \texttt{tstrait}.
% Is this worth the space? We should cite 1KG project also really.
Finally, to demonstrate that \texttt{tstrait} can also be applied
to ARGs inferred from real data, we simulated a trait with 100 causal
sites for the ARG estimated from chromosome 9 of the 1000 Genomes
project data~\citep{kelleher2019} which has YYY samples
and ZZZ variant sites. This took 0.X seconds and Y GB of RAM.
